Flip the virus:\\ A gametheoretic approach to cybersecurity

### 2015-05-17 IEEE Paper/paper/paper.tex
Formule !!

\usepackage[pdftex]{graphicx}  
\usepackage[cmex10]{amsmath}
\usepackage[dutch]{babel}

\usepackage{float}
\usepackage{caption}
\usepackage[caption = false,subrefformat=parens,labelformat=parens]{subfig}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{textcomp}
\usepackage{gensymb}
\usepackage[titlenumbered,algoruled, linesnumbered]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

@@

A particular kind of frequently occurring threaths are Advanced Persistent Threats (APT). An APT is a targeted cyber attack that targets organisations in a stealthy way and that can stay undetected for a long period. This makes it so hard to protect a network or a system against an APT. Bruce Schneier describes an APT as something different and stronger than a conventional threat: ''\textit{A conventional hacker or criminal isn't interested in any particular target. He wants a thousand credit card numbers for fraud, or to break into an account and turn it into a zombie, or whatever. Security against this sort of attacker is relative; as long as you're more secure than almost everyone else, the attackers will go after other people, not you. An APT is different; it's an attacker who - for whatever reason - wants to attack you. Against this sort of attacker, the absolute level of your security is what's important. It doesn't matter how secure you are compared to your peers; all that matters is whether you're secure enough to keep him out}'' - Bruce Schneier: APT is a Useful Buzzword [].\\

@@

This paper proposes the adaptation of the FlipIt formulas as presented in \cite{FlipIt} such as to take the delay for virus propagation into account. In the next section we first present the original FlipIt game. Then section [] presents the FlipIt game with virus propagation. Section [] presents some related work. Section [] concludes the paper and presents avenues for further research.

@@

The state of the resource is denoted as a time-dependent variable $C=C_{i}(t)$. 
$C_{D}(t)$ is 1 if the game is under control by the defender and 0 if the game is under control by the attacker. Reversely, $C_{A}(t)$ will be 1 if the game is under control by the attacker and 0 if under controle by the defender. So, $C_{A}(t)= 1 - C_{D}(t)$.
The game starts with the defender being in control: $C_{D}(0)= 1$. \\

@@

 \begin{table}
 \centering
 \begin{tabular}{ l | c  }
  \textbf{Categories} & \textbf{Classes of Strategies} \\
  \hline Non-adaptive (NA) & Renewal \\
  &  ~~~~ Periodic \\
  &  ~~~~ Exponential \\
  & General non-adaptive \\
  \hline Adaptive (AD) & Last move (LM) \\
  & Full History (FH) \\  
\end{tabular}
 \caption{Hierarchy of Classes of strategies in FlipIt}
 \label{table:Strategies}
 \end{table}

The study of the different strategies by means of FlipIt framework allows to derive a number of interesting results:  
\begin{itemize}
\item periodic games dominate the other renewal strategies, meaning that it is always advantagous to play periodically against an opponent with a renewal strategy;
\item periodic games are disadvantagous against players following a Last Move adaptive strategy;
\item if the defender plays with a periodic rate that is fast enough he'll force the attacker to drop out;
\item any amount of feedback about the opponent received during the game, benefits to a player.
\end{itemize}

@@

Similarly as in [], we split the formalization in two cases. The first case is where the defender plays at least as fast as the attacker, the second case is where the attacker plays at least as fast as the defender. For each of these cases, first the benefit formula of the basic case without delay is presented, and then the delay is introduced.  \\

@@

Correspondeingly, the benefit for the defender can be expressed as:
\begin{equation}\label{first}
\beta_{D}(\alpha_{D},\alpha_{A}) =1 -  \dfrac {r} {2} - k_{D} \alpha_{D} = 1 - \dfrac {\delta_{D}} {2\delta_{A}} - k_{D} \alpha_{D} 
\end{equation}

However, because of the delay required for virus propagation, the maximal time of control is reduced to $\delta_{D}-d$. There is a probability of \textit{r} that the attacker will move in the interval of the defender. However, the gain will not be half of the interval. Indeed, the attacker has to play soon enough to gain control, meaning that the attacker has to play during the period of $\delta_{D}-d$ during the interval of the defender. The probability that the attacker plays soon enough is $\dfrac{\delta_{D}-d}{\delta_{D}}$ and this will give the attacker an avarage gain of $\dfrac{\delta_{D}-d}{2}$. If the attacker moves after the period of $\delta_{D}-d$, the gain of the attacker will be zero. The probability that this happens is  $\dfrac{d}{\delta_{D}}$. The average gain rate of the attacker can then be expressed as follows if we look at one interval of the defender:
\begin{equation}\label{first}
\gamma_{A}(\alpha_{D},\alpha_{A}) = \dfrac {1}{\delta_{D}} [ \dfrac{\delta_{D}}{\delta_{A}} \cdot \dfrac{\delta_{D}-d}{\delta_{D}} \cdot \dfrac{\delta_{D}-d}{2} + \dfrac{\delta_{D}}{\delta_{A}} \cdot \dfrac{d}{\delta_{D}} \cdot 0 ]
\end{equation}

To derive the benefit, the cost of moving is substracted from the average gain. 
\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) = \dfrac { (\delta_{D}-d) ^{2}} {2 \cdot \delta_{D}  \delta_{A}} - k_{A} \alpha_{A}
\end{equation}
\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) = \dfrac { \delta_{D}} {2 \cdot \delta_{A}} - k_{A} \alpha_{A} - (\dfrac{d}{\delta_{A}} - \dfrac{d^{2}}{2 \cdot \delta_{A} \delta_{D}})
\end{equation}

@@

\subsubsection*{\textbf{Case a:} $d + \delta_{A} \leq \delta_{D}$}
~~\\
Consider a timespan $\delta_{A} + d$, representing the attacker's interval followed by the delay period in his next interval. The defender will never move twice during this timespan because $\delta_{A} + d \leq \delta_{D}$. The defender will move during the interval of the attacker with a probability of $\dfrac{\delta_{A}}{\delta_{D}} $. When this happens the defender will end with being in control at the end of the interval. In the next interval the attacker will have to regain control, meaning that during the delay, the defender stays in control, see figure \ref{fig:case2} cases (1) and (2). This means that the defender will keep the control over the resource in the next interval over a period of the delay, namely \textit{d}. Because $d + \delta_{A} \leq \delta_{D}$ the next move of the defender in this second interval will never occur during the delay, meaning that the entire delay can be considered as an extra benefit resulting of a play in the previous interval. 
So, every time the defender plays, he will get an average gain of $\dfrac{\delta_{A}}{2}$ in the interval where he plays and in the next interval will always receive a extra gain of $d$, yielding a total average gain per interval of
$\dfrac{(d+\dfrac{\delta_{A}2})}{\delta_{A}}$

@@

It is crucial that $ \delta_{D}$ is at least as large as $d + \delta_{A}$. If not, this would mean that the defender can move during the delay in the interval following the interval where the defender already moved. This would mean that there can be an overlap between the average gain of $\dfrac{\delta_{a}}{2}$ and the delay. The above benefit formula would then include to much gain for the defender: the potential overlap during the delay would be counted twice. \\
To obtain the formula in case of a too long delay, we therefore need to substract this overlapping gain from the above formula.  The overlapping gain is equal to 
~~ \\
\subsubsection*{\textbf{Case b:} $d + \delta_{A} > \delta_{D}$}
~~~\\

@@

Another addition of Laszka et al. to the game of FlipIt \cite{MitigationCovert} 
is extending the game to also consider non-targeted attacks by non-strategic players. In this game the defender tries to maintain control over the resource that is subjected to both targeted and non-targeted attacks. Non-targeted attacks can include phishing, while targeted attacks may include threats delivered through zero day attack vulnerabilities. \\
One of the last important additions from Laszka et al. \cite{MitigationNonTargeted} is to consider a game where the moves made by the attacker are still covert but the moves made by the defender are known to the attacker. This means that the attacker can base his attacks on the defender's moves. Both the targeted and non-targeted attacks don't succeed immediately. [verschil uitleggen] For the targeted attack the time till it succeeds is given by an exponential distributed random variable with a known rate. The non-targeted attacks are modelled as a single attacker and the time till it succeeds is given by a Poisson process. The conclusion of this paper is that the optimal strategy for the defender is moving periodically. \\

@@

Finally researchers also have investigated the behaviour of humans playing FlipIt. A. Nochenson and Grossklags \cite{nochenson2013behavioral}  investigate how people really act when given temporal decisions. Reitter et al. \cite{reitter2013risk} observed continuous games, 20-seconds FlipIt game.

\section{Conclusions and Furhter Research}
\label{ch:conclusion}
In this paper we presented an adaptation of the FlipIt game to the situation of virus propagation, such as to take the delay for network infection into account. We discerned two cases. In the case the defender plays faster than the attacker, the attacker simply looses the delay. In the case the attacker plays faster, each time the defender plays in an interval, he will gain the extra time of the delay. The delay is therefore always detrimental to the benefit of the attacker. 
The formula for the second case is a slight overestimation of the defender's benefit in case the defender plays faster than the attacker period extende with the delay. When the defender plays in an interval, he will gain extra time in the next interval, since the attacker has to regain control. However, not the entire delay will be extra time, as the defender may move in again sooner. 
Futher research needs to be performed to calculate the exact overestimation in this case.\\
Also the impact of the delay on nash equilibria and the determination of optimal defender and attacher strategies needs further investigation.

>>>>

\usepackage[pdftex]{graphicx}  
\usepackage[cmex10]{amsmath}

\usepackage{float}
%maakt caption van figuren groter
\usepackage{caption}
\usepackage{amsmath}
\usepackage{epstopdf}
\usepackage{textcomp}
\usepackage{gensymb}
\usepackage[titlenumbered,algoruled, linesnumbered]{algorithm2e}
\usepackage{booktabs}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}

@@

A particular kind of frequently occurring threats are Advanced Persistent Threats (APT). An APT is a targeted cyber attack that targets organisations in a stealthy way and that can stay undetected for a long period. This makes it so hard to protect a network or a system against an APT. Bruce Schneier describes an APT as something different and stronger than a conventional threat: ''\textit{A conventional hacker or criminal isn't interested in any particular target. He wants a thousand credit card numbers for fraud, or to break into an account and turn it into a zombie, or whatever. Security against this sort of attacker is relative; as long as you're more secure than almost everyone else, the attackers will go after other people, not you. An APT is different; it's an attacker who - for whatever reason - wants to attack you. Against this sort of attacker, the absolute level of your security is what's important. It doesn't matter how secure you are compared to your peers; all that matters is whether you're secure enough to keep him out}'' - Bruce Schneier \cite{APTBruce}.\\

@@

This paper proposes the adaptation of the FlipIt formulas as presented in \cite{FlipIt} such as to take the delay for virus propagation into account. In the next section we first present the original FlipIt game. Then section \ref{ch:FlipItGame} presents the FlipIt game with virus propagation. Section \ref{ch:extendedWork} presents some related work. Section \ref{ch:conclusion} concludes the paper and presents avenues for further research.

@@

The state of the resource is denoted as a time-dependent variable $C=C_{i}(t)$. 
$C_{D}(t)$ is 1 if the game is under control by the defender and 0 if the game is under control by the attacker. Reversely, $C_{A}(t)$ will be 1 if the game is under control by the attacker and 0 if under control by the defender. So, $C_{A}(t)= 1 - C_{D}(t)$.
The game starts with the defender being in control: $C_{D}(0)= 1$. \\

@@

 \begin{table}
 \centering
 \begin{tabular}{ l | c  }
  \textbf{Categories} & \textbf{Classes of Strategies} \\
  \hline Non-adaptive (NA) & Renewal \\
  & - Periodic \\
  & ~~~ - Exponential \\
  & General non-adaptive \\
  \hline Adaptive (AD) & Last move (LM) \\
  & Full History (FH) \\  
\end{tabular}
 \caption{Hierarchy of Classes of strategies in FlipIt}
 \label{table:Strategies}
 \end{table}

The study of the different strategies by means of FlipIt framework allows to derive a number of interesting results:  
\begin{itemize}
\item periodic games dominate the other renewal strategies, meaning that it is always advantageous to play periodically against an opponent with a renewal strategy;
\item periodic games are disadvantageous against players following a Last Move adaptive strategy;
\item if the defender plays with a periodic rate that is fast enough he'll force the attacker to drop out;
\item any amount of feedback about the opponent received during the game, benefits to a player.
\end{itemize}

@@

Similarly as in \cite{FlipIt}, we split the formalization in two cases. The first case is where the defender plays at least as fast as the attacker, the second case is where the attacker plays at least as fast as the defender. For each of these cases, first the benefit formula of the basic case without delay is presented, and then the delay is introduced.  \\

@@

Correspondingly, the benefit for the defender can be expressed as:
\begin{equation}\label{first}
\beta_{D}(\alpha_{D},\alpha_{A}) =1 -  \dfrac {r} {2} - k_{D} \alpha_{D} = 1 - \dfrac {\delta_{D}} {2\delta_{A}} - k_{D} \alpha_{D} 
\end{equation}

However, because of the delay required for virus propagation, the maximal time of control is reduced to $\delta_{D}-d$. There is a probability of \textit{r} that the attacker will move in the interval of the defender. However, the gain will not be half of the interval. Indeed, the attacker has to play soon enough to gain control, meaning that the attacker has to play during the period of $\delta_{D}-d$ during the interval of the defender. The probability that the attacker plays soon enough is $\dfrac{\delta_{D}-d}{\delta_{D}}$ and this will give the attacker an average gain of $\dfrac{\delta_{D}-d}{2}$. If the attacker moves after the period of $\delta_{D}-d$, the gain of the attacker will be zero. The probability that this happens is  $\dfrac{d}{\delta_{D}}$. The average gain rate of the attacker can then be expressed as follows if we look at one interval of the defender:
\begin{equation}\label{first}
\gamma_{A}(\alpha_{D},\alpha_{A}) = \dfrac {1}{\delta_{D}} [ \dfrac{\delta_{D}}{\delta_{A}} \cdot \dfrac{\delta_{D}-d}{\delta_{D}} \cdot \dfrac{\delta_{D}-d}{2} + \dfrac{\delta_{D}}{\delta_{A}} \cdot \dfrac{d}{\delta_{D}} \cdot 0 ]
\end{equation}

To derive the benefit, the cost of moving is subtracted from the average gain. 
\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) = \dfrac { (\delta_{D}-d) ^{2}} {2 \cdot \delta_{D}  \delta_{A}} - k_{A} \alpha_{A}
\end{equation}
\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) = \dfrac { \delta_{D}} {2 \cdot \delta_{A}} - k_{A} \alpha_{A} - ( \dfrac{d^{2}}{2 \cdot \delta_{A} \delta_{D}} - \dfrac{d}{\delta_{A}} )
\end{equation}

@@

\subsubsection*{\textbf{Case a:} $d + \delta_{A} \leq \delta_{D}$}
~~\\
Consider a timespan $\delta_{A} + d$, representing the attacker's interval followed by the delay period in his next interval. The defender will never move twice during this timespan because $\delta_{A} + d \leq \delta_{D}$. The defender will move during the interval of the attacker with a probability of $\dfrac{\delta_{A}}{\delta_{D}} $. When this happens the defender will end with being in control at the end of the interval. In the next interval the attacker will have to regain control, meaning that during the delay, the defender stays in control, see figure \ref{fig:case2} cases (1) and (2). This means that the defender will keep the control over the resource in the next interval over a period of the delay, namely \textit{d}. Because $d + \delta_{A} \leq \delta_{D}$ the next move of the defender in this second interval will never occur during the delay, meaning that the entire delay can be considered as an extra benefit resulting of a play in the previous interval. 
So, every time the defender plays, he will get an average gain of $\dfrac{\delta_{A}}{2}$ in the interval where he plays and in the next interval will always receive a extra gain of $d$, yielding a total average gain per interval of
$\dfrac{(d+\dfrac{\delta_{A}}{2})}{\delta_{A}}$

@@

It is crucial that $ \delta_{D}$ is at least as large as $d + \delta_{A}$. If not, this would mean that the defender can move during the delay in the interval following the interval where the defender already moved. This would mean that there can be an overlap between the average gain of $\dfrac{\delta_{A}}{2}$ and the delay. The above benefit formula would then include to much gain for the defender: the potential overlap during the delay would be counted twice. \\

~~ \\
\subsubsection*{\textbf{Case b:} $d + \delta_{A} \geq \delta_{D}$}
~~~\\

To obtain the formula in case of a too long delay, we therefore need to subtract this overlapping gain from the above formula. 
Since $\delta_{D} \geq \delta_{A}$, if the defender enters the interval immediately after the attacker has played, then the defender cannot have played in the previous interval. In that case, there is no overlap. So the problem of the overlap only appears if the defenders enters late enough and thus only the last part of the delay is subject to overlap. The larger the difference between the interval of the defender and the attacker, the smaller the risk of overlap. Concretely, only the last part of lengt $d - (\delta_{D} - \delta_{A})$ is subject to overlap. Hence, the probability of overlap is $\dfrac{ d - (\delta_{D} - \delta_{A})}{\delta_{D}}$ and the gain will be half of this interval:  $\dfrac{ d - (\delta_{D} - \delta_{A})}{2}$.  The gain rate to be subtracted is therefore:\\

\begin{equation}\label{first}
\dfrac{1} {\delta_{A}} \cdot \dfrac{d - (\delta_{D} - \delta_{A})}{\delta_{D}} \cdot \dfrac{d - (\delta_{D} - \delta_{A})}{\delta_{D}}
\end{equation}

The total gain  rate of the defender is obtained by subtracting this term from the gain rate of case a:
 \begin{equation}\label{first}
\gamma_{D}(\alpha_{D},\alpha_{A}) = \dfrac{\delta_{A}}{\delta_{D}} \cdot \dfrac{(d+\dfrac{\delta_{A}}{2})}{\delta_{A}} - \dfrac{(d - (\delta_{D} - \delta_{A}))^{2}}{2 \delta_{D} \delta_{A}}
\end{equation}
\begin{equation}\label{first}
\gamma_{D}(\alpha_{D},\alpha_{A}) = \dfrac{\delta_{A}}{2\delta_{D}} + \dfrac{d}{\delta_{D}} - \dfrac{(d - (\delta_{D} - \delta_{A}))^{2}}{2 \delta_{D} \delta_{A}}
\end{equation}\\
This yields in the following benefit formula:
\begin{equation}\label{first}
\beta_{D}(\alpha_{D},\alpha_{A}) = \dfrac{\delta_{A}}{2\delta_{D}} + \dfrac{d}{\delta_{D}} - k_{D} \alpha_{D} - \dfrac{(d - (\delta_{D} - \delta_{A}))^{2}}{2 \delta_{D} \delta_{A}}
\end{equation}\\

The benefit for the attacker will be as follows:
\begin{equation}\label{first}
\beta_{A}(\alpha_{D},\alpha_{A}) = 1 -\dfrac{\delta_{A}}{2\delta_{D}} - \dfrac{d}{\delta_{D}} - k_{A} \alpha_{A} + \dfrac{(d - (\delta_{D} - \delta_{A}))^{2}}{2 \delta_{D} \delta_{A}}
\end{equation}\\

@@

Another addition of Laszka et al. to the game of FlipIt \cite{MitigationCovert} 
is extending the game to also consider non-targeted attacks by non-strategic players. In this game the defender tries to maintain control over the resource that is subjected to both targeted and non-targeted attacks. Non-targeted attacks can include phishing, while targeted attacks may include threats delivered through zero day attack vulnerabilities. \\
One of the last important additions from Laszka et al. \cite{MitigationNonTargeted} is to consider a game where the moves made by the attacker are still covert but the moves made by the defender are known to the attacker. This means that the attacker can base his attacks on the defender's moves. Both the targeted and non-targeted attacks don't succeed immediately. For the targeted attack the time till it succeeds is given by an exponential distributed random variable with a known rate. The non-targeted attacks are modelled as a single attacker and the time till it succeeds is given by a Poisson process. The conclusion of this paper is that the optimal strategy for the defender is moving periodically. The difference with this paper is that the delay in this paper is dependent on the number of nodes that have to be flipped in a network. This cannot be modelled with the framework in the Laszka et al. paper because the delay is chosen as an exponential distributed random variable. Another difference is that in the case of the Laszka et al. paper, the moves of the defender are considered not stealthy and so the attacker knows when the defender plays. \\

@@

Finally researchers also have investigated the behaviour of humans playing FlipIt. A. Nochenson and Grossklags \cite{nochenson2013behavioral}  investigate how people really act when given temporal decisions. Reitter et al. \cite{reitter2013risk} extended the work of A. Nochenson and Grossklags to include various visual presentation modalities for the available feedback during the investigation.

\section{Conclusions and Further Research}
\label{ch:conclusion}
In this paper we presented an adaptation of the FlipIt game to the situation of virus propagation, such as to take the delay for network infection into account. We discerned two cases. In the case the defender plays faster than the attacker, the attacker simply looses the delay. In the case the attacker plays faster, each time the defender plays in an interval, he will gain the extra time of the delay. The delay is therefore always detrimental to the benefit of the attacker. 
The formula for the second case is a slight overestimation of the defender's benefit in case the defender plays faster than the attacker period extended with the delay. When the defender plays in an interval, he will gain extra time in the next interval, since the attacker has to regain control. However, not the entire delay will be extra time, as the defender may move in again sooner. 
Further research needs to be performed to calculate the exact overestimation in this case.\\
Also the impact of the delay on Nash equilibria and the determination of optimal defender and attacher strategies needs further investigation.
